---
title: "HW_02 - Justin Nelson"
output: github_document
---

```{r setup, include=FALSE}
require(knitr)
knitr::opts_chunk$set(echo = TRUE)
opts_knit$set(root.dir = "/Users/JustinNelson/Desktop/Junior Fall/Stat 133/stat133-hws-fall17/hw02/data")
library(ggplot2)
library(dplyr)
```

```{r}
dat <- read.csv(file="nba2017-player-statistics.csv")
colnames(dat)
position <- factor(c('C', 'PF', 'PG', 'SF', 'SG'))
position
dat <- read.csv(file="nba2017-player-statistics.csv", colClasses = c('character', 'character', "factor", 'character', 'double', rep("integer", 19)))

str(dat)
```

```{r}
library(readr)
data <- read_csv(file="nba2017-player-statistics.csv", col_types = list(Position = col_factor(c('C', 'PF', 'PG', 'SG', 'SF'))))
str(data)
```

```{r}
data$Experience <- gsub('R', 1, data$Experience)
data <- transform(data, Experience = as.integer(Experience))
str(data)
```

```{r}
data$Missed_FG <- data$FGA - data$FGM
data$Missed_FT <- data$FTA - data$FTM
data$PTS <- (data$Points2 * 2) + (data$FTM * 1) + (data$Points3 * 3)
data$REB <- data$OREB + data$DREB
data$MPG <- data$MIN / data$GP
data$EFF <- (data$PTS + data$REB + data$AST + data$STL + data$BLK - data$Missed_FG - data$Missed_FT - data$TO) / data$GP
summary(data$EFF)
hist(data$EFF)

top_10 <- slice(arrange(select(data, Player, Team, Salary, EFF), desc(EFF)), 1:10)
top_10

neg <- select(filter(data, EFF < 0), Player, EFF)
neg

cpts <- cor(data$EFF, data$PTS)
creb <- cor(data$EFF, data$REB)
cast <- cor(data$EFF, data$AST)
cblk <- cor(data$EFF, data$BLK)
cmfg <- -cor(data$EFF, data$Missed_FG)
cmft <- -cor(data$EFF, data$Missed_FT)
cto <- -cor(data$EFF, data$TO)
cgp <- cor(data$EFF, data$GP)

cpts
creb
cast
cblk
cmfg
cmft
cto
cgp

correlations <- sort((c(cpts, creb, cast, cblk, cmfg, cmft, cto)), decreasing = TRUE)
correlations
barplot(correlations, names.arg = c('points', 'rebs', 'asst', 'blocks', 'mfg', 'mft', 'TO'))

```
```{r}
eff_salary_graph <- ggplot(data, aes(x = EFF, y = Salary)) + geom_point() + geom_smooth(method = "loess")
eff_salary_graph

eff_salary_corr <- cor(data$EFF, data$Salary)
eff_salary_corr

#This graph and the correlation tells us that there is a positive, and somewhat strong correlation between EFF and salary - aka, the better a player is as determined by the EFF metric, they get paid more.

players2 <- filter(data, MPG > 20)
players2_graph <- ggplot(players2, aes(x = EFF, y = Salary)) + geom_point() + geom_smooth(method = "loess")
players2_graph

players2_cor <- cor(players2$EFF, players2$Salary)
players2_cor

#To my surprise, there is actually a lower correlation between "established" players EFF and salary. Upon further thought, this makes sense because teams are likely willing to pay high-promising rookies a huge amount of money to lock them in, despite them not proving their worth in the NBA yet. Those who are more established have lower long-term potential due to their age, and also the "hype" around them has probably diminished.
```

Nothing was exceptionally hard.

Grasping ggplot was easy after doing today's lab.

Like homework 1, I referenced online internet forums like stackexchange a lot to help me out with syntax to complete problems in this homework.

About 2.5 hours.

It was all about equally distributed.

Nothing stumped me entirely.

Nothing frustrating in particular.

This analysis and the results that came out of them were actually pretty interesting and not something I expected. I genuinely enjoyed this assignment.
